Designing Types for R, Empirically

### 2020-10-12 paper/main.tex
intro

\begin{abstract}
The R programming language is widely used in a variety of domains.  It
was designed to favor an interactive style of programming with minimal
syntactic and conceptual overhead. This design is well suited to
interactive data analysis, but a bad fit for tools such as compilers
or program analyzers which must generate native code or catch
programming errors.  In particular, R has no type annotations, and all
operations are dynamically checked at run-time. The starting point for
our work are the twin questions: \emph{what expressive power is needed
to accurately type R code?} and \emph{which type system is the R
community willing to adopt?} Both questions are difficult to answer
without actually experimenting with a type system. The goal of this
paper is to provide data that can feed into that design process. To
this end, we perform a large corpus analysis to gain insights in the
degree of polymorphism exhibited by idiomatic R code and explore
potential benefits that the R community could accrue from a simple
type system.  As a starting point, we infer type signatures for
\NUMFUNCTIONS functions from \NUMPACKAGES packages among the most
widely used open source R libraries. We then conduct an evaluation on
\NUMPKGSEVAL clients of these packages, as well as on end-user code
found on the Kaggle competition website.
\end{abstract} \maketitle

@@

Our community builds, improves, and reasons about programming languages. To
make design decisions that benefit users, we need to understand our target
language as well as the real-world needs it answers. Often, we can appeal to
our intuition, as many languages are intended for general purpose
programming tasks.  Unfortunately, intuition may fail when looking at
domain-specific languages designed for a specific
group of users to solve very specific needs. This is the case of the data
science language R.

R and its ancestor S were designed, implemented, and maintained by
statisticians. Originally they aimed to be glue languages for reading data
and calling statistical routines written in Fortran. Over three decades they
became widely used across many fields for data analysis and visualization.
Modern R, as an object of study, is fascinating. It is a vectorized,
dynamically typed, lazy functional language with limited side-effects,
extensive reflective facilities and retrofitted object-oriented programming
support.

Many of the design decisions that gave us R were intended to foster an
interactive and exploratory programming style. These include, to name a few,
the lack of type annotations, the ability to use syntactic shortcuts, and the
automatic conversion between data types. While these choices have led to a
language with a low barrier to entry---many data science educational
programs do not teach R itself but simply introduce some of its key
libraries---they have also created a language where errors can easily go
undetected.

Retrofitting a type system to the R language would increase our assurance in
the result of data analysis, but we are faced with two challenges. First,
it is unclear what would be the \emph{right} type system for a language as
baroque as R. For example, one of the most popular data type, the
\code{data.frame}, is manipulated through reflective operations---a data
frame is a table whose columns can be added or removed on the fly.  Second,
but just as crucially, designing a type system that will be adopted would
require overcoming some prejudices and educating large numbers of users.

>>>>

\begin{abstract}
The R programming language is widely used in a variety of domains.  It
was designed to favor an interactive style of programming with minimal
syntactic and conceptual overhead. This design is well suited to data
analysis, but a bad fit for tools such as compilers or program
analyzers. In particular, R has no type annotations, and all
operations are dynamically checked at run-time. The starting point for
our work are the twin questions: \emph{what expressive power is needed
to accurately type R code?} and \emph{which type system is the R
community willing to adopt?} Both questions are difficult to answer
without actually experimenting with a type system. The goal of this
paper is to provide data that can feed into that design process. To
this end, we perform a large corpus analysis to gain insights in the
degree of polymorphism exhibited by idiomatic R code and explore
potential benefits that the R community could accrue from a simple
type system.  As a starting point, we infer type signatures for
\NUMFUNCTIONS functions from \NUMPACKAGES packages among the most
widely used open source R libraries. We then conduct an evaluation on
\NUMPKGSEVAL clients of these packages, as well as on end-user code
from the Kaggle competition.
\end{abstract} \maketitle

@@

Our community builds, improves, and reasons about programming
languages. To make design decisions that benefit users, we need to
understand our target language as well as the real-world needs it
answers. Often, we can appeal to intuition, as many languages are
intended for general purpose programming tasks.  Unfortunately,
intuition may fail us when looking at domain-specific languages
designed for a particular group of users to solve specific needs. This
is the case of the data science language R.

R and its ancestor S were designed, implemented, and maintained by
statisticians. Originally they aimed to be glue languages for 
statistical routines written in Fortran. Over three decades they
became widely used across many fields for exploratory data analysis.
Modern R, as an object of study, is fascinating. It is a vectorized,
dynamically typed, lazy functional language with limited side-effects,
extensive reflective facilities and retrofitted object-oriented programming
support.

Many of the design decisions that gave us R were intended to foster an
interactive and exploratory programming style. These include, to name
a few salient ones, the lack of type annotations, the ability to use
syntactic shortcuts, and widespread conversion between data
types. While these choices have decreased the barrier to entry---many
data science educational programs do not teach R itself but simply
introduce some of its key libraries---they also allow for errors to go
undetected.

Retrofitting a type system to the R language would increase our assurance in
the result of data analysis, but this requires facing two challenges. First,
it is unclear what would be the \emph{right} type system for a language as
baroque as R. For example, one of the most popular data type, the
\code{data.frame}, is manipulated through reflective operations---a data
frame is a table whose columns can be added or removed on the fly.  Second,
but just as crucially, designing a type system that will be adopted would
require overcoming some prejudices and educating large numbers of users.
